### 2013-10-01 manuscripts/dpp-msbayes/dpp-msbayes.tex
Incorporating more edits of dpp ms from Mark.

\begin{linenomath}
With \alignmentVector, \hkyModelVector, \ploidyScalarVector, and
\mutationRateScalarConstantVector in hand, the joint posterior distribution
of our model is given by
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \alignmentVector, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\alignmentVector)},
    \label{eq:fullModelCompact}
\end{equation}
where
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
is a vector of population divergence times for each of the \npairs{} pairs of
populations,
$\demographicParamVector = (\demographicParams{1}, \ldots,
\demographicParams{\npairs{}})$
is a vector of the demographic parameters for each of the \npairs{} population
pairs,
$\locusMutationRateScalarVector = (\locusMutationRateScalar{1}, \ldots
\locusMutationRateScalar{\nlociTotal})$
is a vector of locus-specific mutation-rate scaling parameters for each of the
\nlociTotal loci,
\locusRateHetShapeParameter is the shape parameter of a gamma-distributed
prior on \locusMutationRateScalar{}, and
$p(\alignmentVector)$ is the probability of the data (or the marginal
likelihood of the model).
\end{linenomath}

@@

\subsubsection*{Flexibility in parameterizing the model}
We also provide the user with the ability to control the richness of
the model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

We also allow the user to eliminate the parameters associated with
the post-divergence bottlenecks in the descendant populations of
each pair
(\bottleTime{}, \bottleScalar{1}{}, and \bottleScalar{2}{}),
which constrains the descendant populations to be of
constant size from present back to the divergence event.
Also, rather than eliminate the bottleneck parameters, the user
can constrain \bottleScalar{1}{} and \bottleScalar{2}{} to be
equal to remove one free parameter from the model for each of
the population pairs.

Overall, our implementation allows the user to specify a model that has as many
as seven parameters per population pair
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{},
\bottleTime{}, \bottleScalar{1}{}, \bottleScalar{2}{}, and
\migrationRate{})
or as few as one parameter per pair
(\myTheta{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescence-time parameters.

@@

\subsubsection*{Ordering of taxon-specific summary statistics}
As eluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, they, along with the observed summary statistics, are re-ordered by
descending values of average pairwise differences between the descendant
populations \citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeMapVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

Thus, the theoretical basis for this reshuffling is questionable.
It may be justifiable as an additional coarsening of the data,
by removing the information associated with the identity of the
population pairs.
However, we do not provide a formal proof that this re-ordering of the pairs
does not introduce bias.
One can imagine situations in which the sampling intensity (i.e., the
number of gene copies, loci, and locus length) is highly skewed across
pairs and/or there is large heterogeneity in mutational processes
(e.g., \hkyModel{}{}, \mutationRateScalarConstant{}{}) among the pairs.
In such cases it seems possible that discarding this information could
cause bias.
Furthermore, given that part of the motivation for re-ordering by $\pi_b$ is to
minimize Euclidean distances between simulated datasets in which the true model
has a single divergence event, and this reordering increases the model's
tendency to infer a single divergence event \citep{Huang2011}, it seems
possible that this approach could be biasing the method \citep{Oaks2012}.

@@

\subsection*{Validation analyses: Ordered divergence models}
Our results show that the model estimating ordered models of divergence
\modelDPPOrdered performs similarly to its counterpart sampling over
unordered models (\modelDPP) when the prior is correct (Figs.\
\labelcref{figValidationAccuracyDPPOrderedDPPOrdered,figValidationModelChoiceDPPOrderedDPPOrdered}).

>>>>

\begin{linenomath}
With \alignmentVector, \hkyModelVector, \ploidyScalarVector, and
\mutationRateScalarConstantVector in hand, the joint posterior distribution
of our model is given by
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \alignmentVector, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\alignmentVector \given \geneTreeVector, \divTimeMapVector,
        \demographicParamVector, \locusMutationRateScalarVector,
        \locusRateHetShapeParameter, \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\geneTreeVector, \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
        \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)}{
        p(\alignmentVector)},
    \label{eq:fullModelCompactJoint}
\end{equation}
which can be expanded out into components that are assumed to be independent
to get
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \alignmentVector, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\alignmentVector)},
    \label{eq:fullModelCompact}
\end{equation}
where
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
is a vector of population divergence times for each of the \npairs{} pairs of
populations,
$\demographicParamVector = (\demographicParams{1}, \ldots,
\demographicParams{\npairs{}})$
is a vector of the demographic parameters for each of the \npairs{} population
pairs,
$\locusMutationRateScalarVector = (\locusMutationRateScalar{1}, \ldots
\locusMutationRateScalar{\nlociTotal})$
is a vector of locus-specific mutation-rate scaling parameters for each of the
\nlociTotal loci,
\locusRateHetShapeParameter is the shape parameter of a gamma-distributed
prior on \locusMutationRateScalar{}, and
$p(\alignmentVector)$ is the probability of the data (or the marginal
likelihood of the model).
\end{linenomath}

@@

\subsubsection*{Flexibility in parameterizing the model}
In our implementation, we provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

We also provide the option of eliminating the parameters associated with the
post-divergence bottlenecks in the descendant populations of each pair
(\bottleTime{}, \bottleScalar{1}{}, and \bottleScalar{2}{}),
which constrains the descendant populations to be of
constant size from present back to the divergence event.
Also, rather than eliminate the bottleneck parameters,
we allow \bottleScalar{1}{} and \bottleScalar{2}{} to be constrained to be
equal, which removes one free parameter from the model for each of the
population pairs.

Overall, our implementation allows an investigator to specify a model that has
as many as seven parameters per population pair
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{},
\bottleTime{}, \bottleScalar{1}{}, \bottleScalar{2}{}, and
\migrationRate{})
or as few as one parameter per pair
(\myTheta{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescence-time parameters.

@@

\subsubsection*{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeMapVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

Thus, the theoretical basis for this reshuffling is questionable.
It may be justifiable as an additional coarsening of the data,
by removing the information associated with the identity of the
population pairs.
However, we do not provide a formal proof that this re-ordering of the pairs
does not introduce bias.
One can imagine situations in which the sampling intensity (i.e., the
number of gene copies, loci, and locus length) is highly skewed across
pairs and/or there is large heterogeneity in mutational processes
(e.g., \hkyModel{}{}, \mutationRateScalarConstant{}{}) among the pairs.
In such cases, it seems possible that discarding this information could
cause bias.
Furthermore, given that (1) part of the motivation for re-ordering by $\pi_b$
is to minimize Euclidean distances between simulated datasets in which the true
model has a single divergence event, and (2) the reordering increases the
model's tendency to infer a single divergence event \citep{Huang2011}, it seems
possible that this approach could be biasing the method \citep{Oaks2012}.

@@

\subsection*{Validation analyses: Ordered divergence models}
Our results show that the ordered models of divergence, \modelDPPOrdered,
perform similarly to sampling over unordered models (\modelDPP) when the prior
is correct (Figs.\
\labelcref{figValidationAccuracyDPPOrderedDPPOrdered,figValidationModelChoiceDPPOrderedDPPOrdered}).
This suggests that the method is not adversely affected by the large increase
in possible discrete models (from 22 unordered to 4140 ordered models) when
there are 8 pairs of populations.
